---
title: "Chapter 6"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of longitudinal data

Data wrangling exercises are done in the corresponding R script.

## Analysis exercises

### Part 1. RATS Data

Firstly, the data is read from csv-file:
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

RATS <- read.csv("data/RATSL.csv")

RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

```
Now let's draw the plot:

```{r}
ggplot(RATS, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS$Weight), max(RATS$Weight)))
```


There are 8 rats in Group 1 and 4 rats in Group 2 and Group 3. For Group 1 and Group 3 the weigth seems to lie in the certain range: for Group 1: from 225 to almost 300; for Group 3: from about 460 to 560. However, there is an outlier in the Group 2: one rat was 500 gr at week 0 and gained about 80 gr. Same for other 2 rats from Group 2, while the weight of the 3rd remained didn't change so much.
It is also worth noting that there are patterns in gaining weight: e.g. in Group 3 there is a decrease which is present for all 4 rats.

Let's draw the plot for the same data, but standardised:

```{r}
RATS <- RATS %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

ggplot(RATS, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "right") + 
  scale_y_continuous(name='standardized weight')
```

Group 1 is below 0, near -1, Group 2 is slightly above 0 (except for the outlier) and Group 3 is near 1.


Let's look at average profiles for each group:
```{r}
n <- RATS$Time %>% unique() %>% length()

RATS1 <- RATS %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = (sd(Weight))/sqrt(n)) %>%
  ungroup()

ggplot(RATS1, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```


The patterns of weight changing are similar for Group 2 and Group 3, e.g. on the 45rd day there is a decrease in weight, but already on the next day it has risen and continues rising. It is also worth noting that the weight in Group 2 is more diverse and the least difference between weight is in Group 1.

```{r}
#Now let's do the analysis of covariance: 
#t.test rises an error
RATS2 <- RATS %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

#RATS%>%t.test(mean ~ Group)

RATS1 <- read.csv("data/RATS.csv")

RATS2 <- RATS2 %>%
  mutate(baseline = RATS1$WD1)

fit <- lm(mean ~ baseline + Group, data = RATS2)
anova(fit)
```

### Part 2. BPRS Data

Firstly, the data is read from csv-file:
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

BPRS <- read.csv("data/BPRSL.csv")

BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
```

Now let's draw a plot of the data:
```{r}
ggplot(BPRS, aes(x=subject, y=bprs, color=treatment)) + geom_point() + geom_smooth()
```


Creating a regression model:

```{r}
BPRS_reg <- lm(bprs ~ week + treatment, BPRS)
summary(BPRS_reg)
```

The model shows that the average BPRS is 46.45 and that with every week the BPRS dropped by 2 points on average, but the treatment 2 leaded to its increase by 0.57. However, the t-value of treatment to is near to zero, so we must accept the null hypothesis that there is no relationship between treatment 2 and bprs. On the other side, the situation is opposite for the week variable: it seems that that we can reject the null hypothesis because of t-value and Pr(>|t|), so there is a relationship between two variables.

Creating a random intercept model:
```{r}
library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRS, REML = FALSE)
summary(BPRS_ref)
```

Creating a random intercept and random slope model:

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRS, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)
```

Creating a model with interaction:

```{r}
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + (week*treatment), data = BPRS, REML = FALSE)

summary(BPRS_ref2)

anova(BPRS_ref2, BPRS_ref1)
```